Locally Conservative Discontinuous Petrov-Galerkin Finite Elements for Fluid Problems

### 2013-11-18 docs/Truman/LocalConservation/paper/paper.tex
Added Stokes and Burgers results to local conservation paper.

\subsection{Importance of Local Conservation}
Locally conservative methods hold a special place for numerical analysts in
the field of fluid dynamics.
Perot\cite{Perot2011} argues
\begin{quote}
Accuracy, stability, and consistency are the mathematical concepts that are
typically used to analyze numerical methods for partial differential equations
(PDEs). These important tools quantify how well the mathematics of a PDE is
represented, but they fail to say anything about how well the physics of the
system is represented by a particular numerical method. In practice, physical
fidelity of a numerical solution can be just as important (perhaps even more
important to a physicist) as these more traditional mathematical concepts. A
numerical solution that violates the underlying physics (destroying mass or
entropy, for example) is in many respects just as flawed as an unstable
solution.
\end{quote}

@@

\subsection{Inviscid Burgers' Equation}
The inviscid Burger's equation is
\[
\frac{\partial u}{\partial t}+u\frac{\partial u}{\partial x}=f\,.
\]
Define the space-time gradient: $\Grad_{xt}=\LRp{\frac{\partial}{\partial
x},\frac{\partial}{\partial t}}^T$. We can now rewrite this as
\[
\vecttwo{u}{1}\cdot\Grad_{xt}u=0\,.
\]
Multiplying by a test function $v$, and integrating by parts:
\[
-\LRp{\vecttwo{u}{1}u,\Grad_{xt}v}+\LRa{\hat t,v}=\LRp{f,v}\,,
\]
where $\hat t$ is the trace of $\vecttwo{u^2}{u}\cdot\mathbf{n}_{xt}$ on
element boundaries, and $\mathbf{n}_{xt}$ is the space-time normal vector.
As in convection-diffusion, local conservation implies that
$\int_{\partial K}\hat t=\int_Kf$ for all elements, $K$.

In order to solve this nonlinear problem, we linearize and do a simple Newton iteration
until the solution converges. The results follow in Section \ref{sec:inviscidBurgers}.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{figs/Erickson/graphFlux.pdf}
\caption{Flux imbalance in Erickson-Johnson solutions}
\label{ericksonFlux}
\end{figure}

@@

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/Discontinuous/graph8nc.png}
\caption{Nonconservative}
\label{fig:discontinuousModified8nc}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/Discontinuous/robust8nc.png}
\caption{Conservative}
\label{fig:discontinuousModified8c}
\end{subfigure}
\caption{Discontinuous source problem after 8 refinements}
\label{fig:discontinuous}
\end{figure}

@@

\subsubsection{Inviscid Burgers' Equation}\label{sec:inviscidBurgers}

\subsubsection{Stokes Flow Around a Cylinder}\label{sec:stokesCylinder}
This is a common problem used to stress-test local conservation properties of
least squares finite element methods. Since DPG can be viewed as a generalized
least squares methods\cite{DPGOverview}, we might expect it to struggle with
this problem as well. The problem domain is detailed in
Figure~\ref{fig:stokesCylinderDomain} with inlet and outlet velocity profiles
\[
\bfu_{in}=\bfu_{out}=\vecttwo{(1-y)(1+y)}{0}\,,
\]
and zero flow on the cylinder and at the top and bottom walls. We use $\mu=$
with both Stokes problems.

We measure mass loss much more directly in these two Stokes problems. Because
fluid enters and leaves the domain only through the inlet and outlet
boundaries, we should be able to integrate the mass flux over any
cross-section of the mesh and get the same value. Unfortunately, it's not
mathematically kosher to take line integrals of our field variables which only
live in $L^2$. We can, however, integrate the flux over element boundaries.
This carries the unfortunate limitation that we can only measure mass loss
where there is a clear vertical mesh line. We therefore pick integration lines
from the initial coarse mesh and measure the mass flux after each adaptive refinement
step. The percent mass loss is thus
\[
\%m_{loss}=\frac{\int_{\Gamma_{in}}\bfu\cdot\bfn_{in}d\ell-\int_S\bfu\cdot\bfn_Sd\ell}
{\int_{\Gamma_{in}}\bfu\cdot\bfn_{in}d\ell}\times100.
\]

@@

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag9_NC0.png}
\caption{Nonconservative on initial mesh}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag9_C0.png}
\caption{Conservative on initial mesh}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag9_NC6.png}
\caption{Nonconservative after 6 refinement steps}
\label{fig:stokesCylinderNC}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag9_C6.png}
\caption{Conservative after 6 refinement steps}
\label{fig:stokesCylinderC}
\end{subfigure}
\caption{Stokes flow around a cylinder - velocity magnitude}
\label{fig:stokesCylinder}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{figs/StokesCylinder/MassLoss9_NC.pdf}
\caption{Mass loss in Stokes flow around a cylinder - nonconservative DPG}
\label{fig:stokesCylinderMassLossNC}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{figs/StokesCylinder/MassLoss9_C.pdf}
\caption{Mass loss in Stokes flow around a cylinder - conservative DPG}
\label{fig:stokesCylinderMassLossC}
\end{figure}

\subsubsection{Stokes Flow Over a Backward Facing Step}\label{sec:stokesStep}
Similarly, least squares methods have historically performed very poorly when
calculating Stokes flow over a backward facing step. The stress singularity at
the reentrant corner, in particular, seems to destroy local conservation. We
assign parabolic inlet and outlet velocity boundary conditions
\[
\bfu_{in}=\vecttwo{8(y-0.5)(1-y)}{0}
\quad\text{ and }\quad
\bfu_{out}=\vecttwo{y(1-y)}{0}
\]
and zero velocity on all other boundaries.

@@

\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{figs/StokesStep/MassLoss_NC.pdf}
\caption{Mass loss in Stokes backward facing step - nonconservative DPG}
\label{fig:stokesStepMassLossNC}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{figs/StokesStep/MassLoss_C.pdf}
\caption{Mass loss in Stokes backward facing step - conservative DPG}
\label{fig:stokesStepMassLossC}
\end{figure}

\subsection{Analysis of Results}\label{sec:problemAnalysis}
% Solution quality
The general trend we see looking at the results is that the solution quality
of the conservative and nonconservative formulations is nearly identical. This
is ideal, as we wish to preserve the attractive features of DPG --
namely that it minimizes error in the energy norm. In particular, when we
examine the error convergence plots for the Erickson-Johnson problem, the
energy levels for the two methods lie nearly right on top of each other. This
seems to indicate at the least that enforcing local conservation doesn't hurt
DPG too much.

% Overshoots/undershoots
Another measure of solution quality is to look at the magnitude of overshoots
and undershoots in the solution. In most of the problems considered (barring
the discontinuous source problem) the exact solution would range exactly
between 0 and 1. In the resolved solutions, the divergence from these values
would be negligible, but for the inner layer, in which $\epsilon=10^{-6}$, we
get overshoots and undershoots along the separation line. For the
nonconservative formulation, the solution range for $u$ was $[-0.392,1.342]$,
while the conservative solution range was $[0.389,1.378]$. This difference is
easily accounted for by the differences in refinement patterns between the two
methods.

@@

\section{Conclusions}
We've developed a locally conservative formulation of DPG by leveraging
Lagrange multipliers. While the formulation is provably convergent, it does
increase the number of unkowns in the system and changes the structure from
symmetric positive-definite to a saddlepoint problem. Numerical results
indicate that the method delivers what it promises with local flux imbalances
hovering close to machine precision, but they also indicate that for the
problems considered, standard DPG is close to conservative and becomes more so
with refinements. Probably the most encouraging result of these experiments is
that enforcing local conservation did not change the nature of the solutions
too significantly. Standard DPG has a lot of attractive features, and we wish
to preserve thoes.

Mass conservation is probably the foremost issue facing the class of Least
Squares Finite Element Methods. And there are several benchmark problems in the
literature where these methods have performed very poorly, particularly Stokes
flow over a cylinder and the backward facing step problem. Extending our locally
conservative formulation of DPG to these kinds of problems should be
relatively straightforward. The real test of whether local conservation is
worth it for DPG will come when we start doing more realistic problems.

>>>>

\subsection{Importance of Local Conservation}
Locally conservative methods hold a special place for numerical analysts in
the field of fluid dynamics.
Perot\cite{Perot2011} argues
\begin{quote}
Accuracy, stability, and consistency are the mathematical concepts that are
typically used to analyze numerical methods for partial differential equations
(PDEs). These important tools quantify how well the mathematics of a PDE is
represented, but they fail to say anything about how well the physics of the
system is represented by a particular numerical method. In practice, physical
fidelity of a numerical solution can be just as important (perhaps even more
important to a physicist) as these more traditional mathematical concepts. A
numerical solution that violates the underlying physics (destroying mass or
entropy, for example) is in many respects just as flawed as an unstable
solution.
\end{quote}
There are also some mathematically attractive reasons to pursue local
conservation. The Lax-Wendroff theorem guarantees that a convergent numerical
solution to a system of hyperbolic conservation laws will converge to the
correct weak solution.

@@

\subsection{Inviscid Burgers' Equation}
We include the inviscid Burgers' equation in our suite of tests because, being
a nonlinear hyperbolic conservation law, it
falls under the purvue of the Lax-Wendroff theorem.
The inviscid Burger's equation is
\[
\frac{\partial u}{\partial t}+u\frac{\partial u}{\partial x}=f\,.
\]
Define the space-time gradient: $\Grad_{xt}=\LRp{\frac{\partial}{\partial
x},\frac{\partial}{\partial t}}^T$. We can now rewrite this as
\[
\Grad_{xt}\cdot\vecttwo{u^2/2}{u}=0\,.
\]
Multiplying by a test function $v$, and integrating by parts:
\[
-\LRp{\vecttwo{u^2/2}{u},\Grad_{xt}v}+\LRa{\hat t,v}=\LRp{f,v}\,,
\]
where $\hat t$ is the trace of $\vecttwo{u^2/2}{u}\cdot\mathbf{n}_{xt}$ on
element boundaries, and $\mathbf{n}_{xt}$ is the space-time normal vector.
As in convection-diffusion, local conservation implies that
$\int_{\partial K}\hat t=\int_Kf$ for all elements, $K$.

In order to solve this nonlinear problem, we linearize and do a simple Newton iteration
until the solution converges. The linearized equation is
\[
-\LRp{\vecttwo{u}{1}\Delta u,\Grad_{xt}v}+\LRa{\hat t,v}=\LRp{f,v}
+\LRp{\vecttwo{u^2/2}{u},\Grad_{xt}v} \,,
\]
where $u$ is the previous solution iteration and $\Delta u$ is the update.
The results follow in Section \ref{sec:inviscidBurgers}.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{figs/Erickson/modifiedFlux.pdf}
\caption{Flux imbalance in Erickson-Johnson solutions}
\label{ericksonFlux}
\end{figure}

@@

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/Discontinuous/modified8nc.png}
\caption{Nonconservative}
\label{fig:discontinuousModified8nc}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/Discontinuous/modified8c.png}
\caption{Conservative}
\label{fig:discontinuousModified8c}
\end{subfigure}
\caption{Discontinuous source problem after 8 refinements}
\label{fig:discontinuous}
\end{figure}

@@

\subsubsection{Inviscid Burgers' Equation}\label{sec:inviscidBurgers}
This is a standard test problem for Burgers' equation. The domain is a unit
square. We assign boundary conditions $\hat t=-(1-2x)$ on the bottom, $\hat
t=-1/2$ on the left, while $\hat t=1/2$ on the right. Since this is a
hyperbolic equation, there is no need to set a boundary condition on the top.

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/Burgers/graph8nc.png}
\caption{Nonconservative}
\label{fig:burgers8nc}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/Burgers/graph8c.png}
\caption{Conservative}
\label{fig:burgers8c}
\end{subfigure}
\caption{Burgers' problem after 8 refinements}
\label{fig:burgers}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=0.8\textwidth]{figs/Burgers/graphFlux.pdf}
\caption{Flux imbalance in Burgers' solutions}
\label{fig:burgers_flux}
\end{figure}

\subsubsection{Stokes Flow Around a Cylinder}\label{sec:stokesCylinder}
This is a common problem used to stress-test local conservation properties of
least squares finite element methods. Since DPG can be viewed as a generalized
least squares methods\cite{DPGOverview}, we might expect it to struggle with
this problem as well. The problem domain is detailed in
Figure~\ref{fig:stokesCylinderDomain} with inlet and outlet velocity profiles
\[
\bfu_{in}=\bfu_{out}=\vecttwo{(1-y)(1+y)}{0}\,,
\]
and zero flow on the cylinder and at the top and bottom walls. We use $\mu=$
with both Stokes problems and set velocity boundary conditions on $\hat\bfu$.

Bochev \etal~\cite{Bochev2010} run this test with both $r=0.6$ and $r=0.9$; we
repeat the same experiments with standard and locally conservative DPG methods
starting from the very coarse meshes shown in Figure~\ref{fig:stokesMesh}
while adaptively refining toward a resolved solution. The extreme pressure
gradient in the $r=0.9$ case obviously makes local conservation more
challenging.

We measure mass loss much more directly in these two Stokes problems. Because
fluid enters and leaves the domain only through the inlet and outlet
boundaries, we should be able to integrate the mass flux over any
cross-section of the mesh and get the same value. Unfortunately, it's not
mathematically kosher to take line integrals of our field variables which only
live in $L^2$. We can, however, integrate the trace and flux variables over element boundaries.
This carries the unfortunate limitation that we can only measure mass loss
where there is a clear vertical mesh line. We therefore pick integration lines
from the initial coarse mesh and measure the mass flux after each adaptive refinement
step. The percent mass loss is thus
\[
\%m_{loss}=\frac{\int_{\Gamma_{in}}\bfu\cdot\bfn_{in}d\ell-\int_S\bfu\cdot\bfn_Sd\ell}
{\int_{\Gamma_{in}}\bfu\cdot\bfn_{in}d\ell}\times100,
\]
where $S$ is some vertical mesh line.

@@

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.6\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/mesh6_0.png}
\caption{Mesh for $r=0.6$}
\label{fig:stokesMesh6}
\end{subfigure}
\begin{subfigure}[t]{0.6\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/mesh9_0.png}
\caption{Mesh for $r=0.9$}
\label{fig:stokesMesh9}
\end{subfigure}
\caption{Initial mesh for Stokes flow over a cylinder}
\label{fig:stokesMesh}
\end{figure}

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag6_NC6.png}
\caption{Nonconservative with $r=0.6$}
\label{fig:stokesCylinderNC6}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag6_C6.png}
\caption{Conservative with $r=0.6$}
\label{fig:stokesCylinderC6}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag9_NC6.png}
\caption{Nonconservative with $r=0.9$}
\label{fig:stokesCylinderNC9}
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/StokesCylinder/umag9_C6.png}
\caption{Conservative with $r=0.9$}
\label{fig:stokesCylinderC9}
\end{subfigure}
\caption{Stokes flow around a cylinder after 6 refinement steps - velocity magnitude}
\label{fig:stokesCylinder9}
\end{figure}

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.8\textwidth}
\includegraphics[width=\textwidth]{figs/StokesCylinder/MassLoss6_NC.pdf}
\caption{Nonconservative}
\label{fig:stokesCylinder6MassLossNC}
\end{subfigure}
\begin{subfigure}[t]{0.8\textwidth}
\includegraphics[width=\textwidth]{figs/StokesCylinder/MassLoss6_C.pdf}
\caption{Conservative}
\label{fig:stokesCylinder6MassLossC}
\end{subfigure}
\caption{Mass loss in Stokes flow around a cylinder of radius $0.6$}
\label{fig:stokesCylinder6MassLoss}
\end{figure}

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.8\textwidth}
\includegraphics[width=\textwidth]{figs/StokesCylinder/MassLoss9_NC.pdf}
\caption{Nonconservative}
\label{fig:stokesCylinder9MassLossNC}
\end{subfigure}
\begin{subfigure}[t]{0.8\textwidth}
\includegraphics[width=\textwidth]{figs/StokesCylinder/MassLoss9_C.pdf}
\caption{Conservative}
\label{fig:stokesCylinder9MassLossC}
\end{subfigure}
\caption{Mass loss in Stokes flow around a cylinder of radius $0.9$}
\label{fig:stokesCylinder9MassLoss}
\end{figure}

\subsubsection{Stokes Flow Over a Backward Facing Step}\label{sec:stokesStep}
Similarly, least squares methods have historically performed very poorly when
calculating Stokes flow over a backward facing step. The stress singularity at
the reentrant corner, in particular, seems to destroy local conservation. We
assign parabolic inlet and outlet velocity boundary conditions
\[
\bfu_{in}=\vecttwo{8(y-0.5)(1-y)}{0}
\quad\text{ and }\quad
\bfu_{out}=\vecttwo{y(1-y)}{0}
\]
and zero velocity on all other boundaries. In this problem, we solve with
fourth order field and flux variables, fifth order traces, and sixth order
test functions.

@@

\begin{figure}[p]
\centering
\begin{subfigure}[t]{0.8\textwidth}
\includegraphics[width=\textwidth]{figs/StokesStep/MassLoss_NC.pdf}
\caption{Nonconservative DPG}
\label{fig:stokesStepMassLossNC}
\end{subfigure}
\begin{subfigure}[t]{0.8\textwidth}
\includegraphics[width=\textwidth]{figs/StokesStep/MassLoss_C.pdf}
\caption{Conservative DPG}
\label{fig:stokesStepMassLossC}
\end{subfigure}
\caption{Mass loss in Stokes backward facing step}
\label{fig:stokesStepMassLoss}
\end{figure}

\subsection{Analysis of Results}\label{sec:problemAnalysis}
\subsubsection{Convection-Diffusion Results}
% Solution quality
The general trend we see looking at the results is that the solution quality
of the conservative and nonconservative formulations is nearly identical once
sufficiently resolved. This
is ideal, as we wish to preserve the attractive features of DPG --
namely that it minimizes error in the energy norm. In particular, when we
examine the error convergence plots for the Erickson-Johnson problem, the
energy levels for the two methods lie nearly right on top of each other. This
seems to indicate at the least that enforcing local conservation doesn't hurt
DPG too much.

% Overshoots/undershoots
Another measure of solution quality is to look at the magnitude of overshoots
and undershoots in the solution. In most of the convection-diffusion problems considered (barring
the discontinuous source problem) the exact solution would range exactly
between 0 and 1. In the resolved solutions, the divergence from these values
would be negligible, but for the inner layer, in which $\epsilon=10^{-6}$, we
get overshoots and undershoots along the separation line. For the
nonconservative formulation, the solution range for $u$ was $[-0.392,1.342]$,
while the conservative solution range was $[-0.389,1.378]$. This difference is
easily accounted for by the differences in refinement patterns between the two
methods.

@@

\subsubsection{Burgers' Results}
Standard and locally conservative DPG perform nearly identically for the
inviscid Burgers' problem. It's obvious that the Lax-Wendroff condition of
local conservation is a sufficient, but not necessary condition for numerical
solutions to hyperbolic conservation laws. We see the same behavior with the
flux imbalance plots that was so common with convection-diffusion.

\subsubsection{Stokes Results}
The two Stokes problems are the first ones we encounter that really stress the local
conservation property of standard DPG. With a cylinder radius of $0.6$,
standard DPG loses nearly 30\% of the mass post-cylinder, but quickly recovers
most of that with further refinement. As we increase the cylinder radius to
$0.9$, the problem only exacerbates. Nearly 100\% of the mass is lost in the
constricted region on coarse meshes. It takes a much higher level of
resolution to recover the mass loss. The stress singularity at the reentrant
corner of the backward facing step causes issues for standard DPG on coarse
meshes. It seems that the error in approximating the singularity outweighs the
error of missed mass conservation. If we focus refinements at the singularity,
the error eventually drops far enough for the method to become nearly
conservative.
The small amount of mass loss for the conservative method is clearly due to
accumulation of floating point error.

\section{Conclusions}
We've developed a locally conservative formulation of DPG by leveraging
Lagrange multipliers. While the formulation is provably convergent and robust, it does
increase the number of unkowns in the system and changes the structure from
symmetric positive-definite to a saddlepoint problem. Numerical results
indicate that the method delivers what it promises with local flux imbalances
hovering close to machine precision, but they also indicate that for most of the
problems considered, standard DPG is close to conservative and becomes more so
with refinements. For the Stokes problems, standard DPG suffered similar mass
loss as standard least squares on coarse meshes, but made up the lost mass with further
resolution.
Probably the most encouraging result of these experiments is
that enforcing local conservation did not change the nature of the solutions
too significantly. Standard DPG has a lot of attractive features, and we wish
to preserve those.

% Mass conservation is probably the foremost issue facing the class of Least
% Squares Finite Element Methods. And there are several benchmark problems in the
% literature where these methods have performed very poorly, particularly Stokes
% flow over a cylinder and the backward facing step problem. Extending our locally
% conservative formulation of DPG to these kinds of problems should be
% relatively straightforward. The real test of whether local conservation is
% worth it for DPG will come when we start doing more realistic problems.
